RationalGRL: A Framework for Argumentation and Goal Modeling

### 2018-09-14 Response to reviewers2.txt

<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>
<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<< REPLIES FOR REVIEW #2 >>>>>><<<<>>>>>>>>>>>>>>>>>>>>>>
<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>

===== COMMENT #1
In the course of the revision, several of my suggestions for improvement have simply been refused. Hence, I do not consider the improvements sufficient.

===== REPLY #1
We have reconsidered all of your suggestions for improvements in the first revision, and did another cycle of improvements of our paper based on your suggestions. We have replied to those suggestions below.

===== COMMENT #2
To the contrary, the newly added Section 7 is unacceptable from a scientific point of view. First, it is called a "small user evaluation". Then it has a subsection entitled "Experimental setup", but there are no hypotheses. Instead, there is another subsection entitled "Analysis", which starts with "Although our empirical evaluation ...". And why is there another subsection about usability of the tool? If that was the concern, a usability test or a heuristic evaluation would have had to be done.

===== REPLY #2
— Please add here Floris —

<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>
<<<<<<<<<<<<<<<<<<<<<<< REVISION OF REPLIES FOR REBUTTAL #1 >>>>>>>>>>>>>>>>>>>>>>>>>>>>
<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>>

@@

===== ORIGINAL COMMENT #3
I think that the paper should compare the proposed argumentation structure at least with the beliefs in GRL. In general, how does this argumentation approach really differ from one based on beliefs and, in particular, belief revision? Assume that an argument is a belief of some stakeholder revised by another argument, which is a belief of some other stakeholder?

===== ORIGINAL REPLY #3
The connection between argumentation and belief revision is interesting, especially from a theoretical point of view, and in knowledge representation and reasoning this is a popular topic of research (see references 5-7 below). However, since studying relations between argumentation and belief revision does not seem directly related to our current framework and seems more interesting from a formal logical point of view, we have left this out of the discussion in our paper, instead choosing to focus on a comparison with beliefs in GRL.

====== REPLY
In our first revision, we already added a brief comparison between arguments in RationalGRL and beliefs in GRL in the related work section (under "Beliefs in GRL"). We have moved these paragraphs to “Belief Revision“ in section 8.2 (Future work), and further extended and sharpened our discussion and comparison with GRL beliefs. Our main argument, which is now fully in the article, is that in GRL, beliefs are more of an afterthought: they have no formal semantics and it is not possible to further reason with them (i.e. given indicated conflicts between beliefs compute ways in which beliefs can be revised). Furthermore, we briefly discuss belief revision and the possibilities this may offer for future research. A more detailed comparison between RationalGRL and a framework based on belief revision would require us to develop a new belief revision framework connected to GRL and then model, for example, the case study or an extended example using this new belief revision framework for GRL. This would be interesting material for a new paper

@@

===== COMMENT #4 & #5
How would you compare this specific approach with the old gIBIS?
Please discuss why you are trying to have detailed, while simple, rationale, although the recent trend in DR is otherwise?

===== ORIGINAL REPLY #4 & #5
We have added some discussion in the 2nd paragraph of section 8.1-Design Rationale. A brief comparison with, for example gIBIS, is already in section 8.1-Design Rationale (reference [38] includes gIBIS). We felt that an extensive comparison between the ideas behind goal modelling, which we extend with RationalGRL, and design rationale diagramming methods like gIBIS is not within the extent of this paper, where the focus is more on the connection between "normal" goal modelling (GRL) and extended "argumentative" goal modelling (RationalGRL).

@@

===== COMMENT #6
I wonder about the formalization in propositional logic. Why is it not used for automated reasoning (at least partly instead of the specific algorithms)?

===== ORIGINAL REPLY #6
Propositional logic in itself does not contain specific algorithms for, for example, INTRO, REPLACE and DISABLE. Furthermore, the algorithms also introduce arguments and attacks, formal notions of which are not in basic propositional logic. The idea is that the formalization is needed for the algorithms, but that just providing a formal semantics of GRL does not directly tell you how to compute/reason with GRL models.

===== REPLY #6
We have added a paragraph to the introduction of section 5, where we explain that we are in fact doing automatic reasoning by implementing the argumentation semantics, but that, from a practical point of view, we have chosen to provide algorithms instead of a deduction system.

@@

===== COMMENT #7
I am also missing a discussion about the consequences of this binary "nature" of argumentation with respect to argumentation in reality.

===== ORIGINAL REPLY #7
While there is extensive work on less "binary" frameworks for argumentation (see e.g below references 2 and 3), we feel that a discussion on this would not be suitable for the current paper (as it is more of an AI subject than relevant for RE). For an extensive discussion on the relation between the formal argumentation proposed in our article and argumentation in reality, we refer the reviewer to reference [4] below. Note that we have opted not to include these references in the paper, as we are already at 50 references and want to focus more on an RE/SE audience. If the reviewer desires, however, we can add the references to the article.

===== REPLY #7

### 2018-09-14 discussion.tex
fixed some compiling errors
revisions for reviewer 2

\paragraph{Goal Modeling} Argumentation has already been included implicitly in existing goal modeling languages. At its core, goal modeling is a process of argumentation: high-level goals are reasons for lower-level goals, tasks and resources. Hence, refinement and decomposition techniques used in requirements engineering \cite{van2001goal} can be seen as explicit argumentation steps in the goal modeling process. Take, for example, CQ2 of PRAS (Section~\ref{sect:background:pras}), which asks whether there are alternative ways of realizing the same goal. Providing an alternative subgoal or task in a goal model is then an explicit argumentative move in the discussion. In this sense, a goal model provides a justification for itself. This idea is also prevalent in our RationalGRL framework: many arguments are in fact GRL elements, and many critical questions can be answered by introducing new GRL elements. However, as was already argued in Section~\ref{sect:introduction} (and also by \cite{Jureta:RE2008}), argumentation produces different, richer and complementary information to the goal model. The goal model is the product of a process of argumentation, and does not include, for example, goals and tasks that were at some point considered but discarded. Furthermore, for goal models, it is only possible to determine the satisfiability of goals given the possible tasks and resources; what cannot be determined is the acceptability of goals, that is, whether they are \emph{acceptable} (cf. Definition 14} given potentially contradictory opinions of stakeholders.

@@

Considering beliefs, an interesting avenue for further research would be to look at to connections between argumentation and belief revision \cite{alchourron1985logic}, and how these can play a role in RationalGRL and requirements engineering in general. While there is formal work in Artificial Intelligence on the connections between argumentation and belief revision (see e.g. \cite{falappa2009belief}, there is very little work on applying the combination to requirements engineering. The only work we are aware of that deals with his subject is by Gervasi and Zowghi \cite{gervasi2005reasoning}, who use a logical framework for belief revision to detect inconsistencies in software specifications and then revise those specifications so that the inconsistencies are removed. They do not explicitly connect their framework to goal models of goal modelling, however, and do not connect their framework to specific argumentation using schemes in the way RationalGRL does.

@@

>>>>

\paragraph{Goal Modeling} Argumentation has already been included implicitly in existing goal modeling languages. At its core, goal modeling is a process of argumentation: high-level goals are reasons for lower-level goals, tasks and resources. Hence, refinement and decomposition techniques used in requirements engineering \cite{van2001goal} can be seen as explicit argumentation steps in the goal modeling process. Take, for example, CQ2 of PRAS (Section~\ref{sect:background:pras}), which asks whether there are alternative ways of realizing the same goal. Providing an alternative subgoal or task in a goal model is then an explicit argumentative move in the discussion. In this sense, a goal model provides a justification for itself. This idea is also prevalent in our RationalGRL framework: many arguments are in fact GRL elements, and many critical questions can be answered by introducing new GRL elements. However, as was already argued in Section~\ref{sect:introduction} (and also by \cite{Jureta:RE2008}), argumentation produces different, richer and complementary information to the goal model. The goal model is the product of a process of argumentation, and does not include, for example, goals and tasks that were at some point considered but discarded. Furthermore, for goal models, it is only possible to determine the satisfiability of goals given the possible tasks and resources; what cannot be determined is the acceptability of goals, that is, whether they are \emph{acceptable} (cf. Definition 14 given potentially contradictory opinions of stakeholders).

@@

@@

\paragraph{Belief revision} An interesting avenue for further investigation is the theoretical foundation of RationalGRL. Since it is one of our main goals to develop a framework that can be used by practitioners, we developed a formalism that is geared towards implementation using algorithms and a compact propositional logic. However, there is a long strand of research in Artificial Intelligence focusing on developing postulates that characterize how an agent can revise its knowledge base in a rational way~\cite{falappa2009belief}. The most famous postulates are the so-called AGM postulates~\cite{alchourron1985logic}. It would be very interesting to study whether (a variation of) these rationality postulates can be used to characterize the algorithms we have developed in this article, and more generally, what rationality means in our setting. While there is formal work in Artificial Intelligence on the connections between argumentation and belief revision (see e.g. \cite{falappa2009belief}, there is very little work on applying the combination to requirements engineering. The only work we are aware of that deals with his subject is by Gervasi and Zowghi \cite{gervasi2005reasoning}, who use a logical framework for belief revision to detect inconsistencies in software specifications and then revise those specifications so that the inconsistencies are removed. They do not explicitly connect their framework to goal models of goal modeling, however, and do not connect their framework to specific argumentation using schemes in the way RationalGRL does.

### 2018-09-14 formalframework.tex
revisions for reviewer 2

>>>>

We use formal logic (propositional logic) to specify our framework, and in order to develop our tool we have implemented an algorithm to compute argumentation extensions based on this formalization. However, we do not use a dedicated theorem prover or a logical inference engine. The reason is purely practical, it is our aim to develop a tool that practitioners can use in their browser, and it does not seem necessary nor useful to use an external theorem prover for this. Instead, in this section we propose algorithms based on our logical specification, which form the basis of our prototype implementation.

### 2018-09-14 validation.tex
revisions for reviewer 2

\begin{table*}[t]
\centering
\begin{tabularx}{0.95\textwidth}{l|l|l|l|l|l|l|l}
& very difficult & difficult & neither difficult nor easy & easy & very easy & avg. score (1-5) & std. dev.\\
\hline
Q1 & 2 (12.5\%) & 1 (6.3\%) & 6 (37.5\%) & 6 (37.5\% & 1 (6.3\%) & 3.2 & 1.1\\
Q2 & 2 (12.5\%) & 2 (12.5\%) & 4 (25\%) & 8 (50\%) & 0 (0\%) & 3.1 & 1.1\\
Q3 & 0 (0\%) & 3 (20\%) & 4 (26.7\%) & 5 (33.3\%) & 3 (20\%) & 3.5 & 1.1\\
Q4 & 2 (12.5\%) & 5 (31.3\%) & 6 (37.5\%) & 3 (18.8\%) & 0 (0\%) & 2.6 & 1.0
\end{tabularx}
\caption{User scores for the usability of the RationalGRL tool}
\label{table:survey:table1}
\end{table*}

\begin{table*}[t]
\centering
\begin{tabularx}{0.95\textwidth}{l|l|l|l|l|l|l|l}
& very useless & useless & neither useful nor useless & useful & very useful & avg. score (1-5) & std. dev.\\
\hline
Q5 & 0 (0\%) & 0 (0.0\%) & 3 (21.4\%) & 8 (57.1\% & 3 (21.4\%) & 4.0 & 0.7\\
Q6 & 0 (0.0\%) & 3 (21.4\%) & 3 (21.4\%) & 6 (42.9\%) & 2 (14.3\%) & 3.5 & 1.0\\
Q7 & 0 (0\%) & 0 (0\%) & 2 (14.3\%) & 7 (50.0\%) & 5 (35.7\%) & 4.2 & 0.7
\end{tabularx}
\caption{Participant ratings of the usefulness of the additions of RationalGRL}
\label{table:survey:table2}
\end{table*}

\begin{table*}[t]
\centering
\begin{tabularx}{0.95\textwidth}{l|l|l|l|l|l|l|l}
& much more difficult & more difficult & neither more difficult nor easier & easier & much easier & avg. score (1-5) & std. dev.\\
\hline
Q8 & 0 (0\%) & 0 (0.0\%) & 3 (21.4\%) & 8 (57.1\% & 3 (21.4\%) & 4.0 & 0.7\\
Q9 & 0 (0.0\%) & 3 (21.4\%) & 3 (21.4\%) & 6 (42.9\%) & 2 (14.3\%) & 3.5 & 1.0\\
Q10 & 0 (0\%) & 0 (0\%) & 2 (14.3\%) & 7 (50.0\%) & 5 (35.7\%) & 4.2 & 0.7
\end{tabularx}
\caption{Participant ratings of whether the additions of RationalGRL make reasoning about a goal model easier **MARC PLEASE FILL IN THE CORRECT NUMBERS FROM THE SURVEY HERE, TNX}
\label{table:survey:table3}
\end{table*}

First, the participants had to perform the modeling task. They produced RationalGRL models containing on average 3 arguments (in addition  to goals, tasks, etc.) from the short transcript. Some of the examples of RationalGRL models the users created can be found in Appendix~\ref{sect:survey-screenshots}.

After the modelling task, the survey questions were asked. Table~\ref{table:survey:table1} shows the respondents' answers, the average score, and the standard deviation for questions Q1-Q4. 
We also asked the users three open questions related to usability. When asked about the strengths and weaknesses of the RationalGRL tool, users where generally positive about RationalGRL's clear UI, the fact that the tools gives a nice overview of the goals and opinions in a case and that it was straightforward to use and understand. With respect to weaknesses and improvements to the RationalGRL tool, the users mentioned additional UI functionalities such as the possibility to save models, having an ``undo'' function, and flipping the arrow after adding them. Users also suggested various language-related improvements. Some users mentioned they missed the possibility to attack links, while others mentioned that not all GRL elements are supported (for instance, it is currently not possible to add actors).

@@

Tables~\ref{table:survey:table2} and \ref{table:survey:table3} show the respondents' answers, the average score, and the standard deviation for questions Q5-Q7. We concluded with two open questions about the comparison between RationalGRL and other goal modeling languages. When asked about the advantages of RationalGRL over standard goal modeling languages, many users agreed that making arguments explicit may force end users to have a more structured discussion: ``	Clear communication about argumentation and forcing people to think in those clear terms.'', ``...you can add arguments and that you can answer questions that help you to develop arguments'', ``It's useful that discussion and explanation are close to the diagrams'', ``a way to see how decisions are being shaped.''. Furthermore, one user stated that RationalGRL successfully ``tries to capture the rationale behind the modeling process' and ``has a simple way to compute the status of the arguments''. When asked about the weaknesses, users mentioned the complexity as the most important weakness: ``The apparent increase in complexity might lead to negative perceptions'', ``adding yet another layer of complexity scares me''. One user mentioned that this may be a problem with goal modeling in general: ``Goal models are already complex (...) I have worked for years on the effect of context on goal models, and my conclusion is that this was very interesting academic work but with close-to-zero practical implications, unfortunately.''.

@@

For questions Q5 and Q7, the effect sizes are very large\footnote{Where am low p-value indicates whether the fact that average rating of a question is statistically significantly higher than the median 3, the effect size indicates the magnitude of the difference between the measured average and the median. Generally, an effect size below 0.2 is seen as low, and effect size of 0.5 is seen as medium, an effect size of 0.8 is seen as large and an effect size of 1.2 is seen as very large.}. This means that the addition of arguments and the use of arguments to determine the acceptability of elements of the goal model are deemed to be useful to very useful by the participants. This can also be seen in Table~\ref{table:survey:table2}: no participant found the addition of arguments and argument status useless or very useless, and 78.6\% (Q5) respectively 85.7\% (Q7) found the additions useful or very useful. Furthermore, these findings fit with the answers to the open questions (see Section \ref{sec:survey:results}). Finally, the statistical power for these questions is very high (>0.99), which means that the chances of wrongly rejecting H1$_{0}$ are close to 0.

@@

\paragraph{Hypothesis H3}
\begin{figure}[t]
\centering
\includegraphics[width=\textwidth]{img/Table_analysisQ1-Q4.png}
\caption{Statistical analysis of the results of Q1-Q4 **MARC CAN YOU CONVERT THIS INTO A PROPER LATEX TABLE? THANKS!}
\label{fig:statsanalysis2}
\end{figure}

>>>>

\begin{table*}[t]
\centering
\begin{tabularx}{0.95\textwidth}{l|l|l|l|l|l||l|l|l|l|l}
& very difficult & difficult & neutral & easy & very easy & avg (1-5) & std. dev. & p value & power & effect size\\
\hline
Q1 & 2 (12.5\%) & 1 (6.3\%)  & 6 (37.5\%) & 6 (37.5\%) & 1 (6.3\%)  & 3.2 & 1.1 & 0.509 & ?? & ??\\
Q2 & 2 (12.5\%) & 2 (12.5\%) & 4 (25.0\%) & 8 (50.0\%) & 0 (0.0\%)  & 3.1 & 1.1 & 0.652 & ?? & ??\\
Q3 & 0 (0.0\%)  & 3 (20.0\%) & 4 (26.7\%) & 5 (33.3\%) & 3 (20.0\%) & 3.5 & 1.1 & 0.072 & ?? & ??\\
Q4 & 2 (12.5\%) & 5 (31.3\%) & 6 (37.5\%) & 3 (18.8\%) & 0 (0.0\%)  & 2.6 & 1.0 & 0.138 & ?? & ??
\end{tabularx}
\caption{User scores for the usability of the RationalGRL tool}
\label{table:survey:table1}
\end{table*}

\begin{table*}[t]
\centering
\begin{tabularx}{0.95\textwidth}{l|l|l|l|l|l||l|l|l|l|l}
& very useless & useless & neutral & useful & very useful & avg (1-5) & std. dev. & p value & power & effect size\\
\hline
Q5 & 0 (0.0\%) & 0 (0.0\%)  & 3 (21.4\%) & 8 (57.1\%) & 3 (21.4\%) & 4.0 & 0.7 & 0.000 & 0.999 & 1.5 \\
Q6 & 0 (0.0\%) & 3 (21.4\%) & 3 (21.4\%) & 6 (42.9\%) & 2 (14.3\%) & 3.5 & 1.0 & 0.045 & 0.525 & 0.5\\
Q7 & 0 (0.0\%) & 0 (0.0\%)  & 2 (14.3\%) & 7 (50.0\%) & 5 (35.7\%) & 4.2 & 0.7 & 0.000 & 1.000 & 1.7
\end{tabularx}
\caption{Participant ratings of the usefulness of the additions of RationalGRL}
\label{table:survey:table2}
\end{table*}

\begin{table*}[t]
\centering
\begin{tabularx}{0.95\textwidth}{l|l|l|l|l|l||l|l|l|l|l}
& very difficult & difficult & neutral & easy & very easy & avg (1-5) & std. dev. & p value & power & effect size\\
\hline
Q8  & 0 (0.0\%) & 0 (0.0\%) & 4 (30.8\%) & 4 (30.8\%) & 5 (38.5\%) & 4.1 & 0.9 & 0.000 & 0.991 & 1.2\\
Q9  & 0 (0.0\%) & 0 (0.0\%) & 6 (46.2\%) & 5 (38.5\%) & 2 (15.4\%) & 3.7 & 0.8 & 0.003 & 0.925 & 0.9\\
Q10 & 0 (0.0\%) & 1 (8.3\%) & 4 (33.3\%) & 7 (58.3\%) & 0 (0.0\%)  & 3.5 & 0.7 & 0.013 & 0.772 & 0.7
\end{tabularx}
\caption{Participant ratings of whether the additions of RationalGRL make reasoning about a goal model easier}
\label{table:survey:table3}
\end{table*}

First, the participants had to perform the modeling task. They produced RationalGRL models containing on average 3 arguments (in addition  to goals, tasks, etc.) from the short transcript. Some of the examples of RationalGRL models the users created can be found in Appendix~\ref{sect:survey-screenshots}. The full results of the survey can be downloaded from \url{www.rationalgrl.com}, on the page ``Empirical study''.

After the modelling task, the survey questions were asked. Table~\ref{table:survey:table1} shows the respondents' answers, the average score, and the standard deviation for questions Q1-Q4. 
\todo{Marc}{Floris}{Wellicht kun je hier het best even uitleggen wat p value, power, en effect size betekenen?}

We also asked the users three open questions related to usability. When asked about the strengths and weaknesses of the RationalGRL tool, users where generally positive about RationalGRL's clear UI, the fact that the tools gives a nice overview of the goals and opinions in a case and that it was straightforward to use and understand. With respect to weaknesses and improvements to the RationalGRL tool, the users mentioned additional UI functionalities such as the possibility to save models, having an ``undo'' function, and flipping the arrow after adding them. Users also suggested various language-related improvements. Some users mentioned they missed the possibility to attack links, while others mentioned that not all GRL elements are supported (for instance, it is currently not possible to add actors).

@@

Table~\ref{table:survey:table2} shows the respondents' answers, the average score, the standard deviation, p-value, power, and effect size for questions Q5-Q7, and table~\ref{table:survey:table3} shows the results for questions Q8-Q10. We concluded with two open questions about the comparison between RationalGRL and other goal modeling languages. When asked about the advantages of RationalGRL over standard goal modeling languages, many users agreed that making arguments explicit may force end users to have a more structured discussion: ``	Clear communication about argumentation and forcing people to think in those clear terms.'', ``...you can add arguments and that you can answer questions that help you to develop arguments'', ``It's useful that discussion and explanation are close to the diagrams'', ``a way to see how decisions are being shaped.''. Furthermore, one user stated that RationalGRL successfully ``tries to capture the rationale behind the modeling process' and ``has a simple way to compute the status of the arguments''. When asked about the weaknesses, users mentioned the complexity as the most important weakness: ``The apparent increase in complexity might lead to negative perceptions'', ``adding yet another layer of complexity scares me''. One user mentioned that this may be a problem with goal modeling in general: ``Goal models are already complex (...) I have worked for years on the effect of context on goal models, and my conclusion is that this was very interesting academic work but with close-to-zero practical implications, unfortunately.''.

@@

For questions Q5 and Q7, the effect sizes are very large\footnote{Where am low p-value indicates whether the fact that average rating of a question is statistically significantly higher than the median 3, the effect size indicates the magnitude of the difference between the measured average and the median. Generally, an effect size below 0.2 is seen as low, and effect size of 0.5 is seen as medium, an effect size of 0.8 is seen as large and an effect size of 1.2 is seen as very large.}. This means that the addition of arguments and the use of arguments to determine the acceptability of elements of the goal model are deemed to be useful to very useful by the participants. This can also be seen in Table~\ref{table:survey:table2}: no participant found the addition of arguments and argument status useless or very useless, and 78.6\% (Q5) respectively 85.7\% (Q7) found the additions useful or very useful. Furthermore, these findings fit with the answers to the open questions (see Section \ref{sec:survey:results}). Finally, the statistical power for these questions is very high (\textgreater 0.99), which means that the chances of wrongly rejecting H1$_{0}$ are close to 0.

@@
